---
title: "Simulation example  "
subtitle: "Multinomial logit with Differentiated Bertrand competition"
author: "Eviatar Guttman"
date: "Last updated: `r format(Sys.time(), '%d-%m-%Y')`"  
output: 
  html_document: 
    toc: yes
    toc_depth: 2
    toc_float: yes
    toc_collapsed: yes
    number_sections: yes
    smooth_scroll: yes
    fig_caption: yes
    highlight: tango
    theme: flatly
    css: "style.css"
urlcolor: blue
editor_options: 
  chunk_output_type: console
---  

```{r setup, include = F}
# global chunk options
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE,
                      tidy=TRUE,
                      fig.path='Figures_RMD/',
                      fig.align = 'center')

# don't show NA's in tables
options(knitr.kable.NA = '')

```

```{r loading packages etc., include = F}
# load packages
library(IAplots)
library(tidyverse)
library(kableExtra)
# change settings
theme_set(IA_theme())
options(scipen = 999,digits = 3)

# functions
# 1. make nice tables
IA_kable <- function(x){ 
  kable(x)%>% 
    kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width = F) 
} 
# 2. for large tables we have a scroll box 
IA_kable_s <- function(x, wi = 800, hi = 200){
  wi2 <- paste0(wi, "px")
  hi2 <- paste0(hi, "px")
  IA_kable(x) %>% scroll_box(width = wi2, height = hi2 )
}

```


# intro
:::{.blue-box}
In this document I shall demonstrate merger simulation with a synthetic example.
On the supply side assuming a differentiated Bertrand model and on the demand 
side assuming a multinomial logit demand function. 

:::
<br>

The Bertrand system of equations has 4 components:   
(1) Market shares.   
(2) Prices.   
(3) Derivatives of the demand function with respect to the prices.  
(4) Marginal costs of the firms.    



# Pre merger conditions      
For the sake of demonstration the market shares chosen are `0.2, 0.25` and `0.3`.  
prices are `50, 75` and `80` correspondingly.   


```{r, collapse= TRUE}
# 1. market shares
s <- c(0.2, 0.25, 0.3) 
# outside option
(s_0 <- 1 - sum(s) )


# 2. prices
p <- c(50, 75, 80)
```

 __Calculation of derivatives__  
The logit derivative is a function of market shares and the consumer's derivative 
with respect to the price - the model's $alpha$.  
Again, for demonstration, $alpha = 0.1$ is chosen.    
In fact, this is a calibration of the system. One can estimate those parameters 
with a demand estimation (which is beyond the scope of this simple example). 


```{r, collapse= TRUE}
# 3. derivatives
# choose alpha  
alpha <- - 0.1

# derivatives are functions of market shares and alpha
# own derivative
(d_sj_d_pj <- alpha * (1-s) * s)

# cross derivative
(d_sk_d_pj <- - alpha * s %o% s)

# put own derivatives in diagonal
der <- d_sk_d_pj
diag(der) <- d_sj_d_pj

# final result
der



# cleanup
rm(d_sj_d_pj, d_sk_d_pj)

```

__solving the equation systme for the marginal costs__  
As stated above, First order condition has 4 components.    
We have supplied the market shares, the prices and derivatives.
next we solve the system for the marginal costs.   
  
For the solution we create the "owner ship" matrix: 


```{r}
# 4. solving a system of equations for the marginal costs

# ownership matrix
(theta <- matrix( c(1,0,0,0,1,0,0,0,1), nrow = 3))

# solve a system of linear equations
(mc <- solve(theta * der) %*% s + p )
```

We can check the first order condition are valid.

```{r}
# check that F.O.C's are met: 
s + (theta * der) %*% (p - mc)

```

# Merger simulate 
__a merger in the market__   
When a merger happens, the ownership of firms over the product set is changing. 
Here we assume that products 1 and 2 are now under the same ownership. 
For that, we update the ownership matrix. 
The 1st and 2nd elements 1st and 2nd rows of the matrix will equal to 1: 


```{r}
# new ownership matrix: 
(theta_post <- matrix(c(1,1,0,1,1,0,0,0,1), nrow = 3))
```

After changing the ownership matrix, the firm understands that when the price of 
product 1 goes up, some of the customers that choose not to by it anymore will 
land on product 2, hance for those customers, the firm does not loose its sails.
With that knowledge, the firm can raise the price up. The first order condition
is not valid anymore because the old price is not the optimal price. 

```{r}
# FOC don't hold
s + (theta_post * der) %*% p - (theta_post * der) %*% mc
s + (theta_post * der) %*% (p - mc)

```


## Fixed point iteration
Next we demonstrate a fixed point iteration step by step:   
  
There is no analytic solution to this system. we need to use a method of search
to find the new equilibrium that will prevail in the market after the merger. 
assuming convexity and continuity of the system, we can find a solution and know 
it is unique.   
  

according to the model , the average utility from poduct j is: 
$$ \delta_j = x \beta_j - \alpha p + \xi _j $$    
For each product we calculate the average utility.



A change in price will change the utility from the product. 
This in turn, will be reflected in the demand function, the demand decreases when the 
price goes up and vice versa. 

```{r}
# delta = x * b - alpha * p
(delta <- log(s / s_0)) 
exp(delta) / (1 + sum(exp(delta)))
```

## manual iteration
__fixed point first few iterations__

For demonstration, I'll show part of the convergence process stage by stage manually.  
In every iteration 4 stages will take place:   
(1) Solving the firms first order condition to the new prices in relation to the market shares 
of the previous stage.   
(2) Compute the change in the price vector compared to the pre-merger price vector.  
(3) Compute the change in consumers utilities.   
(4) Get the new marked shares out from the demand function.     
  


```{r}
# the price update procedure: 
# 1. solve FOC 1
(p1 <- as.vector(mc + (1/ - alpha) * (1 / (1 -  theta_post %*% s))))

# 2. delta price
(d_p <- (p1 - p))

# 3. delta in utility
(d_delta <- d_p * alpha)

# 4. solve demand for new utility
(s1 <- exp(delta + d_delta) / (1 + sum(exp(delta + d_delta))))
```

get the results and compere to the pre merger status

```{r}
results <-rbind(c(s, p), c(s1, p1)) 
results
```

Repead this process several times.


```{r}
# solve FOC 2:
p2 <- as.vector(mc + (1/ -alpha) * (1/(1 -  theta_post %*% s1)))
d_p <- (p2 - p) 
d_delta <- d_p * alpha
# solve demand 2:
s2 <- exp(delta + d_delta) / (1 + sum(exp( delta + d_delta)))

results <- rbind(results, c(s2, p2))

# solve FOC 3:
p3 <- as.vector(mc + (1/ -alpha) * (1/(1 -  theta_post %*% s2)))
d_p <- (p3 - p) 
d_delta <- d_p * alpha
# solve demand 3:
s3 <- exp(delta + d_delta) / (1 + sum(exp( delta + d_delta)))

results <- rbind(results, c(s3, p3))

# solve FOC 4:
p4 <- as.vector(mc + (1/ -alpha) * (1/(1 -  theta_post %*% s3)))
d_p <- (p4 - p) 
d_delta <- d_p * alpha
# solve demand 3:
s4 <- exp(delta + d_delta) / (1 + sum(exp( delta + d_delta)))

results <- rbind(results, c(s4, p4))


# solve FOC 5:
p5 <- as.vector(mc + (1/ -alpha) * (1/(1 -  theta_post %*% s4)))
d_p <- (p5 - p) 
d_delta <- d_p * alpha
# solve demand 3:
s5 <- exp(delta + d_delta) / (1 + sum(exp( delta + d_delta)))

results <- rbind(results, c(s5, p5))

# solve FOC 6:
p6 <- as.vector(mc + (1/ -alpha) * (1/(1 -  theta_post %*% s5)))
d_p <- (p6 - p) 
d_delta <- d_p * alpha
# solve demand
s6 <- exp(delta + d_delta) / (1 + sum(exp( delta + d_delta)))

results <- rbind(results, c(s6, p6))

# solve FOC 7:
p7 <- as.vector(mc + (1/ -alpha) * (1/(1 -  theta_post %*% s6)))
d_p <- (p7 - p) 
d_delta <- d_p * alpha
# solve demand
s7 <- exp(delta + d_delta) / (1 + sum(exp( delta + d_delta)))

results <- rbind(results, c(s7, p7))

# solve FOC 8:
p8 <- as.vector(mc + (1/ -alpha) * (1/(1 -  theta_post %*% s7)))
d_p <- (p8 - p) 
d_delta <- d_p * alpha
# solve demand
s8 <- exp(delta + d_delta) / (1 + sum(exp( delta + d_delta)))

```


In the next matrix, colums 1-3 are market shares and columns 4-6 are the prices.   
every iteration the jumps in the values decreases.   
```{r}
(results <- rbind(results, c(s8, p8)))

```



## For loop
OK, we got the idea, lets calculate the new equilibrium. 


```{r}
max_iter <- 100
s_in <- s
i <- 0
s_delta_norm <- 1
convergence_matrix <- matrix(nrow = max_iter, ncol = length(s) * 2 + 2)

while(s_delta_norm > 1e-6 & i < max_iter){
  i <- i + 1
  
  # solve F.O.C
  ( p_new <- as.vector(mc + (1/ - alpha ) * ( 1 / (1 -  theta_post %*% s_in) )))
  
  # change in utility
  d_delta <- (p_new - p) * alpha
  
  # solve demand system
  s_new <- exp(delta + d_delta) / (1 + sum(exp(delta + d_delta)))
  
  # norm of change in market shares
  (s_delta_norm <- sqrt(sum((s_in - s_new) ^ 2)))
  
  # save resault for next iteration
  s_in <- s_new
  
  convergence_matrix[i, ] <- c(s_new,  p_new, s_delta_norm, i)
}

colnames(convergence_matrix) <- c("s1", "s2", "s3", "p1", "p2", "p3", "norm", "iteration")
convergence_matrix

```

Compare the data before the merger with the prediction of the simulation about the merger 
effect on prices and market shares

```{r}
# attach preconditions with last row without NA's in the matrix
final_resault <- convergence_matrix[tail(which(rowSums(!is.na(convergence_matrix)) > 0), 1),]

rbind(c(results[1,], NA,0),final_resault)

```  

The market share of the outside option has increased  
```{r}
c(before = s_0, after = 1 - sum(final_resault[1:3]))
```

Present convergence process on a plot  
```{r}
p1 <- data.frame(convergence_matrix) %>% select(iteration, p1, p2, p3) %>% gather(k = "k", v = "prices", 2:4) %>%
  ggplot(aes(x = iteration, y = prices, color = k)) + geom_line(size = 1.2)

p2 <- data.frame(convergence_matrix) %>% select(iteration, s1, s2, s3) %>% gather(k = "k", v = "shares", 2:4) %>%
  ggplot(aes(x = iteration, y = shares, color = k)) + geom_line(size =1.2)

ggpubr::ggarrange(p1, p2, ncol = 1)


```

First order conditions: 

```{r}
s_in + (theta_post*der) %*% (p_new - mc)

```

# Stata comparison
Compare our results with the Stata merger-sim package: 



```{r}
stata_sim <- read.csv("multinomial stata results.csv")
IA_kable(stata_sim)
```




