---
title: "One-level nested logit"
subtitle: "Merger simulation example"
author: "Eviatar Guttman"
date: "Last updated: `r format(Sys.time(), '%d-%m-%Y')`"  
output: 
  html_document: 
    toc: yes
    toc_depth: 2
    toc_float: yes
    toc_collapsed: yes
    number_sections: yes
    smooth_scroll: yes
    fig_caption: yes
    highlight: tango
    theme: flatly
    css: "style.css"
urlcolor: blue
editor_options: 
  chunk_output_type: console
---  

```{r setup, include = F}
# global chunk options
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE,
                      tidy=TRUE,
                      fig.path='Figures_RMD/',
                      fig.align = 'center')
# don't show NA's in tables
options(knitr.kable.NA = '')
```

```{r loading packages etc., include = T}
# load packages
library(tidyverse)
library(magrittr)
options(scipen = 999, digits = 3)
```
# Introduction  
This document shows how to perform merger simulation with assumed One-Level
nested logit demand system.  
To perform the simulation one need to go thorough 4 stages: 

  1.  Prepare the data  
  2. Calculate derivative matrix for the One-level nested logit demand system based on 
    Price, market shares, Alpha and Sigma.  
  3. Solve a system of equations that characterize the first order condition of the
    assumed supply side - Bertrand competition with differentiated products.   
  4. Calculate a post merger new equilibrium based on post merger system of equations.     
  
      
Next, I'll present the 4 functions that perform each of those steps and briefly review some key points.  
A complete explanation needs to be in another document

# Functions

Prepare data set - `f1_init_calibration` creates a new data frame in the 
global environment with necessary variables for the simulation.   
The new data frame's name is `calibration data`.  
  
When calling the function all arguments must to be specified: data frame, prices, quantities,
a vector of firms, vector of nests, market size, the buyer and the seller

```{r}
f1_init_calibration <- function(data, p, q, firm, nest, m_size, buyer, seller){
  
  data <- data %>%
    rename(p = {{p}}, q = {{q}},firm = {{firm}}, nest = {{nest}}, m_size = {{m_size}})
  
  nest_mat <<- outer(data$nest, data$nest, FUN = "==") * 1
  
  calibration_data <<-
    data %>%
    select(p, q, firm, nest, m_size) %>%
    mutate(
      p = as.numeric(p),
      q = as.numeric(q),
      s = q / m_size,
      s_0 = 1 - sum(s),
      s_g = as.numeric(nest_mat %*% s),
      s_jg = s / s_g,
      firm_post = if_else(firm == {{seller}}, {{buyer}}, firm),
      n = n()) %>% na.omit()
}
```
`f2_derivatives`  builds and returns the derivative matrix from the shares, alpha and sigma.    
In one level nested-logit there are 3  types of derivatives:  
(1) Own derivative, (2) cross derivative for products in the same nest, and (3) 
cross derivative for products in a different nest.  
```{r}
f2_derivatives <- function(shares, nest_shares, alpha, sigma){
  
  s     <- shares
  s_jg  <- nest_shares
  n     <- length(s)
  alpha <- - abs(alpha)
  
  # nest derivatives
  cross_different  <- - alpha * s %o% s * (1 - nest_mat)
  cross_same       <- - alpha * (sigma / (1 - sigma) * s_jg + s) %o% s * nest_mat * (1 - diag(x = 1,nrow = n))
  own              <-   alpha * s * (1 / (1-sigma) - sigma / (1-sigma) * s_jg - s)
  
  derivatives      <- cross_different + cross_same + diag(own)
  derivatives
}
```

`f3_pre_merger_conditions` solves the first order condition of the equation system, and calculates
several variables:   
Margin and Lerner index to check profitability, the Delta which is the mean utility
the consumer has with the prevailing prices and a FOC condition to see if its converging to zero.   
The new variable are added the the `calibration_data` in the global environment.  
```{r}
f3_pre_merger_conditions <- function(c_data, derivatives, sigma){
  
  theta <- outer(c_data$firm, c_data$firm, FUN = "==") * 1
  s <- c_data$s
  p <- c_data$p
  
  c_data$mc <- solve(derivatives * theta) %*% s + p
  
  c_data <- c_data %>% mutate(margin = p - mc,
                              lerner_index = (p - mc) / p,
                              FOC = as.vector(s + (theta * derivatives) %*% (p - mc)),
                              delta =  log(s / s_0)  - sigma * log(s_jg)
                              )
  
  calibration_data <<- c_data
}
```

`f4_fixed_point` calculates the new equilibrium with fixed point iteration.    
It uses the `f5_demand`, `f2_derivatives` and solves the FOC iteratively until
solution is reached or until the maximum number of iterations is reached.   
the user has control over the maximum number of iterations and the tolarance desired
for convergence.    
we shall expand about the argument `convergence_factor` later. 
```{r}
f4_fixed_point <- 
  function(data, alpha, sigma, max_iter = 100, convergence_factor = 1, tolerance = 1e-3){
    
    theta_post <- outer(data$firm_post, data$firm_post, FUN = "==") * 1
    s_in       <- data$s
    p          <- data$p
    mc         <- data$mc
    delta      <- data$delta
    der_new    <- f2_derivatives(calibration_data$s, calibration_data$s_jg, alpha, sigma)
    re         <- matrix(nrow = max_iter, ncol = length(s_in) + 2)
    i          <- 0
    s_d_norm   <- 1
    
    while(i < max_iter & s_d_norm > tolerance){
      
      i         <- i + 1
      p_new     <- mc - (solve(der_new * theta_post)) %*% s_in  # new price
      delta_new <- delta - abs(alpha) * as.vector((p_new - p)/ convergence_factor)
      s_new     <- f5_demand(delta_new, sigma)
      der_new   <- f2_derivatives(s_new$s_t, s_new$s_jg_t, alpha, sigma)
      s_d_norm  <- sqrt(sum((s_in - s_new[[1]]) ^ 2)) # measure convergence
      s_in      <- s_new[[1]]    # new price vector to feed in
      re[i,]    <- c(p_new, norm = s_d_norm,iteration = i) # results
      
    }
    
    re       <- data.frame(re) %>% drop_na()
    names(re)<- c(paste("p", 1:length(s_in), sep = "_"), "norm", "iter")
    
    tail     <- tail(re,1)
    
    data$new_prices <- tail %>% select(starts_with("p_")) %>% t() %>% as.vector()
    data$iterations <- tail %>% select(iter) %>% pull()
    data$norm       <- tail %>% select(norm) %>% pull()
    data$new_shares <- s_new[[1]]
    
    calibration_results <<- data %>% select(firm, nest, p, new_prices, s, new_shares,
                                            iterations, norm, q, m_size)
    
    re
  }
```

Every time the prices change, the mean utility of the consumer changes, hence the 
quantity demand will change.   
`f5_demand` calculates the market shares that will prevail given a vector of prices.  
This is the demand function.   
The user need not operate this function. it is called from within `f4_fixed_point`. 
```{r}
f5_demand <- function(delta, sigma){
  
  # demand function
  exponent    <- exp(delta /(1-sigma))
  D_g         <- unique(nest_mat, MARGIN = 1) %*%  exponent
  denominator <- D_g ^ sigma * (1 + sum(D_g^(1-sigma)))
  s_t         <- as.vector(exponent * unique(nest_mat, MARGIN = 2) %*% (1/ as.vector(denominator)))
  
  # nest shares
  s_g  <- as.numeric(nest_mat %*% s_t)  # nest market share
  s_jg <- s_t / s_g                     # share within the nest
  
  data.frame(s_t = s_t, s_jg_t = s_jg, delta_t = delta, n = length(s_t))
}
```

# Build toy data
Create a toy data of 6 products owned by 3 firms, divided into 2 nests.  
market size will be 100. 
```{r}
df <- data.frame(
  company  = c("a",     "a",     "b",    "b", "c", "c"),
  nests    = c("pre",   "pre",   "pre",   2,   2,   2),
  quantity = c( 20,      5,       10,     5,   10,  25),
  price    = c( 60,      40,      50,     45,  30,  30),
  m_size   = 100,
  not_needed_variable = "junk")
df
```

We can prepare the data for simulation using `f1_init_calibration`:   
Note that all argument are needed for the function to know which variable is which:   
prices, quantities, firm, nest, market size, buyer and seller. 
```{r}
f1_init_calibration(df, p = price, q = quantity, firm = company, nest = nests,
                    m_size = m_size, buyer = 'a', seller = 'b')
```

A new df named`calibration_data` was created in the global environment: 
```{r}
calibration_data
```


# Assume regression results  
Next, we need to assume we have alpha and sigma. Either from a demand estimation 
results or from our knowledge of the market. 
```{r}
sigma0.5 <- 0.5
alpha0.1 <- 0.1
```

# Checking Derivatives
We show how the derivative matrix is like for this toy data.  
In actual simulation this function is being called from f4_fixed_point,
so there is no need to call it.   
Arguments need to be called explicitly from within the data: 

```{r}
der <- f2_derivatives(calibration_data$s, calibration_data$s_jg, alpha0.1, sigma0.5)
der
```
The own derivative in the diagonal is always negative while the cross derivatives 
are positive. 


# Pre-merger conditions  
Now we calculate the MC of the firms. all new variables are added to the `calibration_data`  
To solve the system of equations the there no need for sigma. 
Sigma is used to calculate delta
```{r}
f3_pre_merger_conditions(calibration_data, der, sigma0.5)
calibration_data
```

# Just checking the demand system
For the sake of presentation, lets see that the demand function works correctly.  
feeding the delta's consumers had, we should get the same market shares we assumed.  
Every time the price will change, the delta will change and so the shares. 
```{r}
f5_demand(calibration_data$delta, sigma0.5)
```



# Simulate the merger 
Finally, we can simulate the merger.  
Using the fixed point iteration, we get a new df named `calibration_results` with the results of the simulation.  
The function prints the iterated prices in the R console. 
```{r}
f4_fixed_point(calibration_data, alpha0.1, sigma0.5, convergence_factor = 1.2)
calibration_results
```
`calibration_results` gives the new prices and new shares, reports how many iterations were needed 
to converge to the tolerance of 1/1000 and reports the norm of the change in prices in the last iteration.  

Note that when we called `f4_fixed_point` we used a `convergenc_factor` of 1.2.      
As it happens, this data consists of only 6 observations and the fixed point doesn't 
converge naturally.   
At best practice, one should use a `convergence_factor` of 1, and only if
there's a problem to choose a `convergence_factor` > 1.   
Lets see what happens in this example when it is 1: 

```{r}
f4_fixed_point(calibration_data, alpha0.1, sigma0.5, convergence_factor = 1)
```
Note the price and shares jump between two points. 
Also, `calibration_resaults` tells you that the `norm` is far away from zero: 
```{r}
calibration_results
```

